Liveness-Based Garbage Collection for Lazy Languages

### 2016-05-06 paper/paper.tex
updated name
Fixed name issue
Corrected capitalization of headings

% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.
%\cmt{{
%
%\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors
%section.
%
%\author{
% 1st. author
\authorinfo {Prasanna Kumar. K} 
       {IIT Bombay, Mumbai 400076, India}
       {prasannak@cse.iitb.ac.in}
% 2nd. author
\authorinfo {Amitabha Sanyal} 
       {IIT Bombay, Mumbai 400076, India}
       {as@cse.iitb.ac.in}
% 3rd. author
\authorinfo {Amey Karkare}
       {IIT Kanpur, Kanpur 208016, India}
       {karkare@cse.iitk.ac.in}
\maketitle

@@

%------------------------------------------------------------%
\subsection{Motivating example}
\label{sec:motiv}
\input{fig_motiv}   
%%
Figure~\ref{fig:mot-example}   shows  an   example
program and the state of the heap at the program point $\pi$\footnote{We write
$\pi\!:\!e$ to  associate a  label $\pi$ with  the program  point just
before  the  expression  $e$. The label is not part of the syntax.},  i.e., just  before  the  evaluation  of
$(\length\ \pz)$.  The heap is represented  by a graph in which a node
either  represents  atomic  values   ($\NIL$,  integers,  etc.),  or  a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent values of variables or fields.  The figure
shows the  lists \px\  and \pz\  partially evaluated  due to  the
\SIF\ condition 
(\NULLQ~(\CAR~\pz)).
The edges  shown by thick  arrows are those  which are live  at $\pi$.
%% Only those  cell which have a future use  should be preserved
%% during garbage collection, all other
%% cells can be  reclaimed.
Thus, if a GC takes  place at $\pi$ with the heap shown
in Figure~\ref{fig:mot-example}(b),  a liveness-based  collector (LGC)
will preserve only  the cells referenced by $\pz$, and  the live cells
constituting the  closure referenced by $(\CDR~\pz)$.   In contrast, a
reachability-based collector (RGC) will preserve all cells.

@@

\subsection{Organization of the paper}

Section~\ref{sec:defs}  introduces  the   syntax  of  the  programming
language considered  and gives a small-step  operational semantics for
it.  The liveness  analysis for this language and  its soundness proof
is          presented          in          Section~\ref{sec:liveness}.
Section~\ref{sec:computing} describes  the formulation of  liveness as
grammars. We also give a proof  of undecidability of such grammars and
show     how     they     can     be     approximated     by     DFAs.
Section~\ref{sec:GC-scheme}   discusses   details   of   the   garbage
collector, in particular, the use of  liveness DFAs for GC.  We report
our experimental  results in Section~\ref{sec:experiments}  along with
some  observations.  Section~\ref{sec:relatedwork}  discusses previous
work  related  to  GC and  liveness  and  Section~\ref{sec:conclusion}
discusses possible extensions and concludes the paper.
\section{The target language---syntax and semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first-order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective}, where  all actual  parameters to
functions  are  variables.  Whereas  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of closures through the $\LET$ construct.   We
  assume that  $\LET$s  in  our  language  are  non-recursive;  in  the
  expression $\LET\,\, x  \leftarrow s\,\, \IN\,\, e$,  $x$ should not
  occur in  $s$. The restriction of  \LET\ to a single  definition is
for ease  of exposition---generalization to multiple  definitions does
not add conceptual difficulties.  We further restrict each variable in
a program  to be  distinct, so that  no scope  shadowing occurs; this
simplifies reasoning.

@@

\subsection{Liveness analysis for lazy languages}
\label{sec:liveness-analysis}
\input{example-to-illustrate-liveness}

@@

\subsection{Soundness of analysis}

@@

\section{Towards a  computable form of liveness}\label{sec:computing}
%\section{Generating and storing liveness information}\label{sec:computing}
The analysis in Section~\ref{sec:liveness} is fully context-sensitive, 
describing the  liveness sets  in a function body in
terms of a symbolic demand $\sigma$  and \Lfonly. However, we have yet to
describe (i) how to obtain demand transformers \Lfonly\ from the rule
{\sc live-define} , and (ii) how to compute the concrete demand $\sigma$ on
each function. To do so, we first need to modify the liveness rules
to a slightly different form.

%\subsection{Modifying liveness rules}
\paragraph{Symbolic representation of operations:}
The   $\mathit{ref}$   rule   for   \CONS,   shown   in
Figure~\ref{fig:live-judge}, requires us  to remove the
leading  \acar\ and  \acdr\  from the  access paths  in
$\sigma$.  Similarly, the rules  for \CAR, \CDR, \PRIM,
\NULLQ, and \SIF\ require  us to return $\emptyset$, if
$\sigma$      itself       is      $\emptyset$      and
$\lbrace\epsilon\rbrace$  otherwise.  To  realize these
rules  $\sigma$   needs  to  be  known.   This  creates
difficulties  since  we  want to  solve  the  equations
arising from liveness symbolically.

@@

%\subsection{Generating equations for $\Lfonly_{\mathit f}$}
\paragraph{Computing function summaries $\Lfonly_{\mathit f}$:}
Given a  function $\mathit{f}$, we now  describe how to
generate  equations   for  the   demand  transformation
\Lfonly$_\mathit{f}$.        The       program       in
Figure~\ref{fig:mot-example2-a}   serves  as   a  running
example.  Starting with a  symbolic demand $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In
particular, we  consider $\Lanv{x_i}{}$,  the liveness
of the $i^{\mbox{\footnotesize th}}$ parameter $x_i$.   By   the   rule   {\sc
  live-define},   this   should    be   the   same   as
$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:
$$
 \Lf{\length}{1}{\sigma} = \Lanv{\xl}{} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
$$

@@

Note that this equation can also be viewed as a CFG with \{\acdr,
\clazy\} as terminal symbols and \Df{\length}{1} as the sole
non-terminal.
%\subsection{Generating liveness equations \Lv\  for function bodies}
\paragraph{Handling user-defined functions:}
\label{sec:bodylivenessbodies}
To  avoid analyzing  the body  of a  function for  each
call, we calculated the  liveness for the arguments and
the variables in a function  with respect to a symbolic
demand  $\sigma$.   To  get   the  actual  liveness,  we
calculate an  over-approximation of the  actual demands
made by  all the  calls and  calculate the  liveness at
each  GC  point  inside  the  function  based  on  this
approximation.  The {\em summary demand} is
calculated by  taking a union  of the demands  at every
call site of a function analogous to 0-CFA~\cite{Shivers:1988}.

@@

%\subsection{Solving liveness equations---the grammar interpretation}\label{sec:grammar-formulation}      
\paragraph{From liveness sets to context-free grammars:}\label{sec:grammar-formulation}      
The  equations  above can  now  be  re-interpreted as  a  context-free
grammar  (CFG) on  the  alphabet $\lbrace\acar,  \acdr, \bcar,  \bcdr,
\clazy\rbrace$.  Let \var{X} denote  the non-terminal for a variable
$X$ occurring on the LHS of the equations generated from the analysis.
We can  think of  the resulting productions  as being  associated with
several grammars, one for  each non-terminal \var{\Lanv{x}{}} regarded
as a start  symbol. \added{The grammars for  \var{\Df{\length}{1}} and \var{\sigma_{\length}} consist of the following productions}:
%%  As  an example, the grammar for \var{\Lanv{\xl}{}}
%% and \var{\Lanv{\pa}{}} comprise the following productions:
\begin{eqnarray*}
  %% \var{\Lanv{\xl}{}}  &\rightarrow& 
  %% \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\ && \mid
  %% \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\var{\sigma_{\length}}
&\rightarrow&
\epsilon  \mid \clazy\var{\sigma_{\length}}
%% \\
%% \var{\Lanv{\pa}{}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
\end{eqnarray*}
Other  equations  can   be  converted  similarly.   The
language   generated   by  \var{\Lanv{x}{}},   denoted
$\mathscr{L}(\var{\Lanv{x}{}})$,   is    the   desired
solution  of  \Lanv{x}{}.    
Recall that the decision problem that we are interested in
during GC is:

@@

%\vspace*{-6.5mm}
\section{The GC scheme}
\label{sec:GC-scheme}

@@

%\subsection{Garbage collection for references on print stack}
The evaluation of  the top-level expression in a program  is driven by
a printing function~(Section
\ref{sec:semantics}).  
%% Printing of  atomic values is simple  and does
%% not  trigger a  GC.   In case  of  \CONS\ cells,  the
%% \CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
%% \CAR\ and \CDR\  could be closures requiring evaluation,  and this may
%% trigger  more GCs.  When  this happens.   we have  to
%% consider any  references that  might be  on the  print stack  and copy
%% them.
We extend liveness-based  garbage  collection  to
this function.
\section{Experimental evaluation}
\label{sec:experiments}
Our  experimental  setup   consists  of  the  prototypes   of  (a)  an
interpreter  for our  language, (b)  a  liveness analyzer,  and (c)  a
single generation  copying garbage  collector.  The  garbage collector
can be configured  to work on the basis of  reachability (RGC mode) or
use liveness DFAs (LGC mode).
%
Our benchmark consists of programs taken from nofib~\cite{nofib} and
other repositories for functional programs~\cite{PLT-Scheme, gc_bench,
  huffman-sicp}.  We ran the experiments on 8 core
Intel\textsuperscript{\textregistered} Core\textsuperscript{TM}
i7-4770 3.40GHz CPU having 8192KB L2 cache, 16GB RAM, running 64 bit
Ubuntu 14.04.

@@

\vspace*{-2mm}
\section{Related work} 
\label{sec:relatedwork}
The impact of  liveness on the effectiveness of  GC is
investigated in~\cite{Hirzel}.   They observe that  liveness can
significantly  impact   garbage  collection,  but  only   when  it  is
interprocedural.  As far as memory requirement is concerned, our paper
demonstrates this observation.

@@

%% \begin{figure}[t]
%% \centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
%%  \caption{Memory  usage and garbage collection pattern of  {\tt
%% lambda}.}
%% \label{fig:memory-usage-lambda} \figrule
%% \end{figure}
\vspace*{-2mm}
\vskip -5mm
\section{Future work and conclusions}
\label{sec:conclusion}
%% We have  defined a notion of  liveness on structured
%% data; this generalizes classical liveness and strong
%% liveness.    We  started   with   a  general   fully
%% context-sensitive analysis  which we  proved correct
%% with respect  to a minefield semantics  (this models
%% the  effect  of  garbage  collection  between  every
%% evaluation step).

@@

\appendix
\section{Soundness result for liveness analysis under minefield semantics}
\label{sec:sound}
We first need an auxiliary  result about minefield semantics. Consider
a trace of  a minefield execution.  For every  minefield state $(\rho,
S,   \heap,  e(s),   \sigma)$   that   appears  on   the   LHS  of   a
$\rightsquigarrow$ step, the demand $\sigma$ on the expression $e$ (or
application $s$) is  non-null.  This can be proved by  an induction on
the number  of steps leading  to the  minefield state.  The  base step
holds   because  the   demand   $\sigma_{\mathit{all}}$   on  $(\mainpgm)$   is
non-null. For the inductive step we  observe that for each step of the
minefield semantics, if the demand $\sigma$  on the LHS of a minefield
step  is non-null,  the  demand  on the  RHS  is  a transformation  of
$\sigma$ (for example $(\clazy\cup\acar)\sigma$) which is also non-null.

>>>>

% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.
%\cmt{{
%
%\numberofauthors{3} %  in this sample file, there are a *total*
% of EIGHT authors. SIX appear on the 'first-page' (for formatting
% reasons) and the remaining two appear in the \additionalauthors
%section.
%
%\author{
% 1st. author
\authorinfo {Prasanna Kumar. K.} 
       {IIT Bombay, Mumbai 400076, India}
       {prasannak@cse.iitb.ac.in}
% 2nd. author
\authorinfo {Amitabha Sanyal} 
       {IIT Bombay, Mumbai 400076, India}
       {as@cse.iitb.ac.in}
% 3rd. author
\authorinfo {Amey Karkare}
       {IIT Kanpur, Kanpur 208016, India}
       {karkare@cse.iitk.ac.in}
\maketitle

@@

%------------------------------------------------------------%
\subsection{Motivating Example}
\label{sec:motiv}
\input{fig_motiv}   
%%
Figure~\ref{fig:mot-example}   shows  an   example
program and the state of the heap at the program point $\pi$\footnote{We write
$\pi\!:\!e$ to  associate a  label $\pi$ with  the program  point just
before  the  expression  $e$. The label is not part of the syntax.},  i.e., just  before  the  evaluation  of
$(\length\ \pz)$.  The heap is represented  by a graph in which a node
either  represents  atomic  values   ($\NIL$,  integers,  etc.),  or  a
\CONS\  cell  containing  $\CAR$  and  $\CDR$  fields,  or  a  closure
(represented  by  shaded  clouds).   Edges   in  the  graph  are  {\em
  references} and represent values of variables or fields.  The figure
shows the  lists \px\  and \pz\  partially evaluated  due to  the
\SIF\ condition 
(\NULLQ~(\CAR~\pz)).
The edges  shown by thick  arrows are those  which are live  at $\pi$.
%% Only those  cell which have a future use  should be preserved
%% during garbage collection, all other
%% cells can be  reclaimed.
Thus, if a GC takes  place at $\pi$ with the heap shown
in Figure~\ref{fig:mot-example}(b),  a liveness-based  collector (LGC)
will preserve only  the cells referenced by $\pz$, and  the live cells
constituting the  closure referenced by $(\CDR~\pz)$.   In contrast, a
reachability-based collector (RGC) will preserve all cells.

@@

\subsection{Organization of the Paper}

Section~\ref{sec:defs}  introduces  the   syntax  of  the  programming
language considered  and gives a small-step  operational semantics for
it.  The liveness  analysis for this language and  its soundness proof
is          presented          in          Section~\ref{sec:liveness}.
Section~\ref{sec:computing} describes  the formulation of  liveness as
grammars. We also give a proof  of undecidability of such grammars and
show     how     they     can     be     approximated     by     DFAs.
Section~\ref{sec:GC-scheme}   discusses   details   of   the   garbage
collector, in particular, the use of  liveness DFAs for GC.  We report
our experimental  results in Section~\ref{sec:experiments}  along with
some  observations.  Section~\ref{sec:relatedwork}  discusses previous
work  related  to  GC and  liveness  and  Section~\ref{sec:conclusion}
discusses possible extensions and concludes the paper.
\section{The Target Language---Syntax and Semantics}
\label{sec:defs}
Figure~\ref{fig:lang-syntax} describes the syntax  of our language. It
is a first-order language with lazy semantics. Programs are restricted
to        be        in        Administrative        Normal        Form
(ANF)~\cite{chakravarty03perspective}, where  all actual  parameters to
functions  are  variables.  Whereas  this  restriction  does not  affect
expressibility,  this form  has  the benefit  of  making explicit  the
creation of closures through the $\LET$ construct.   We
  assume that  $\LET$s  in  our  language  are  non-recursive;  in  the
  expression $\LET\,\, x  \leftarrow s\,\, \IN\,\, e$,  $x$ should not
  occur in  $s$. The restriction of  \LET\ to a single  definition is
for ease  of exposition---generalization to multiple  definitions does
not add conceptual difficulties.  We further restrict each variable in
a program  to be  distinct, so that  no scope  shadowing occurs; this
simplifies reasoning.

@@

\subsection{Liveness analysis for Lazy Languages}
\label{sec:liveness-analysis}
\input{example-to-illustrate-liveness}

@@

\subsection{Soundness of Analysis}

@@

\section{Towards a  Computable Form of Liveness}\label{sec:computing}
%\section{Generating and storing liveness information}\label{sec:computing}
The analysis in Section~\ref{sec:liveness} is fully context-sensitive, 
describing the  liveness sets  in a function body in
terms of a symbolic demand $\sigma$  and \Lfonly. However, we have yet to
describe (i) how to obtain demand transformers \Lfonly\ from the rule
{\sc live-define} , and (ii) how to compute the concrete demand $\sigma$ on
each function. To do so, we first need to modify the liveness rules
to a slightly different form.

%\subsection{Modifying liveness rules}
\paragraph{Symbolic Representation of Operations:}
The   $\mathit{ref}$   rule   for   \CONS,   shown   in
Figure~\ref{fig:live-judge}, requires us  to remove the
leading  \acar\ and  \acdr\  from the  access paths  in
$\sigma$.  Similarly, the rules  for \CAR, \CDR, \PRIM,
\NULLQ, and \SIF\ require  us to return $\emptyset$, if
$\sigma$      itself       is      $\emptyset$      and
$\lbrace\epsilon\rbrace$  otherwise.  To  realize these
rules  $\sigma$   needs  to  be  known.   This  creates
difficulties  since  we  want to  solve  the  equations
arising from liveness symbolically.

@@

%\subsection{Generating equations for $\Lfonly_{\mathit f}$}
\paragraph{Computing Function Summaries $\Lfonly_{\mathit f}$:}
Given a  function $\mathit{f}$, we now  describe how to
generate  equations   for  the   demand  transformation
\Lfonly$_\mathit{f}$.        The       program       in
Figure~\ref{fig:mot-example2-a}   serves  as   a  running
example.  Starting with a  symbolic demand $\sigma$, we
determine  \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.   In
particular, we  consider $\Lanv{x_i}{}$,  the liveness
of the $i^{\mbox{\footnotesize th}}$ parameter $x_i$.   By   the   rule   {\sc
  live-define},   this   should    be   the   same   as
$\Lf{f}{i}{\sigma}$. Applying this to \length, we have:
$$
 \Lf{\length}{1}{\sigma} = \Lanv{\xl}{} = \clazy\sigma \cup \acdr\Lf{\length}{1}{\clazy\sigma}
  \cup \clazy\Lf{\length}{1}{\clazy\sigma}
$$

@@

Note that this equation can also be viewed as a CFG with \{\acdr,
\clazy\} as terminal symbols and \Df{\length}{1} as the sole
non-terminal.
%\subsection{Generating liveness equations \Lv\  for function bodies}
\paragraph{Handling User-defined Functions:}
\label{sec:bodylivenessbodies}
To  avoid analyzing  the body  of a  function for  each
call, we calculated the  liveness for the arguments and
the variables in a function  with respect to a symbolic
demand  $\sigma$.   To  get   the  actual  liveness,  we
calculate an  over-approximation of the  actual demands
made by  all the  calls and  calculate the  liveness at
each  GC  point  inside  the  function  based  on  this
approximation.  The {\em summary demand} is
calculated by  taking a union  of the demands  at every
call site of a function analogous to 0-CFA~\cite{Shivers:1988}.

@@

%\subsection{Solving liveness equations---the grammar interpretation}\label{sec:grammar-formulation}      
\paragraph{From Liveness Sets to Context-Free Grammars:}\label{sec:grammar-formulation}      
The  equations  above can  now  be  re-interpreted as  a  context-free
grammar  (CFG) on  the  alphabet $\lbrace\acar,  \acdr, \bcar,  \bcdr,
\clazy\rbrace$.  Let \var{X} denote  the non-terminal for a variable
$X$ occurring on the LHS of the equations generated from the analysis.
We can  think of  the resulting productions  as being  associated with
several grammars, one for  each non-terminal \var{\Lanv{x}{}} regarded
as a start  symbol. \added{The grammars for  \var{\Df{\length}{1}} and \var{\sigma_{\length}} consist of the following productions}:
%%  As  an example, the grammar for \var{\Lanv{\xl}{}}
%% and \var{\Lanv{\pa}{}} comprise the following productions:
\begin{eqnarray*}
  %% \var{\Lanv{\xl}{}}  &\rightarrow& 
  %% \clazy\var{\sigma_{\length}} \mid \acdr \var{\Df {\length}{1}}{\clazy\var{\sigma_{\length}}}  \\ && \mid
  %% \clazy\var{\Df{\length}{1}}{\clazy\var{\sigma_{\length}}} \\
  \var{\Df{\length}{1}} &\rightarrow& \clazy \mid
  \acdr\var{\Df{\length}{1}}\clazy
       \mid \clazy\var{\Df{\length}{1}}{\clazy}\\
\var{\sigma_{\length}}
&\rightarrow&
\epsilon  \mid \clazy\var{\sigma_{\length}}
%% \\
%% \var{\Lanv{\pa}{}} &\rightarrow& \clazy \bcar \var{\Df{\length}{1}}
\end{eqnarray*}
Other  equations  can   be  converted  similarly.   The
language   generated   by  \var{\Lanv{x}{}},   denoted
$\mathscr{L}(\var{\Lanv{x}{}})$,   is    the   desired
solution  of  \Lanv{x}{}.    
Recall that the decision problem that we are interested in
during GC is:

@@

%\vspace*{-6.5mm}
\section{The GC Scheme}
\label{sec:GC-scheme}

@@

%\subsection{Garbage collection for references on print stack}
The evaluation of  the top-level expression in a program  is driven by
a printing function~(Section
\ref{sec:semantics}).  
%% Printing of  atomic values is simple  and does
%% not  trigger a  GC.   In case  of  \CONS\ cells,  the
%% \CAR\ is  first printed followed  by \CDR.   In a lazy  language, both
%% \CAR\ and \CDR\  could be closures requiring evaluation,  and this may
%% trigger  more GCs.  When  this happens.   we have  to
%% consider any  references that  might be  on the  print stack  and copy
%% them.
We extend liveness-based  garbage  collection  to
this function.
\section{Experimental Evaluation}
\label{sec:experiments}
Our  experimental  setup   consists  of  the  prototypes   of  (a)  an
interpreter  for our  language, (b)  a  liveness analyzer,  and (c)  a
single generation  copying garbage  collector.  The  garbage collector
can be configured  to work on the basis of  reachability (RGC mode) or
use liveness DFAs (LGC mode).
%
Our benchmark consists of programs taken from nofib~\cite{nofib} and
other repositories for functional programs~\cite{PLT-Scheme, gc_bench,
  huffman-sicp}.  We ran the experiments on 8 core
Intel\textsuperscript{\textregistered} Core\textsuperscript{TM}
i7-4770 3.40GHz CPU having 8192KB L2 cache, 16GB RAM, running 64 bit
Ubuntu 14.04.

@@

\vspace*{-2mm}
\section{Related Work} 
\label{sec:relatedwork}
The impact of  liveness on the effectiveness of  GC is
investigated in~\cite{Hirzel}.   They observe that  liveness can
significantly  impact   garbage  collection,  but  only   when  it  is
interprocedural.  As far as memory requirement is concerned, our paper
demonstrates this observation.

@@

%% \begin{figure}[t]
%% \centerline{\epsfig{file=lambda.eps, height=4cm, width=7cm}}
%%  \caption{Memory  usage and garbage collection pattern of  {\tt
%% lambda}.}
%% \label{fig:memory-usage-lambda} \figrule
%% \end{figure}
\vspace*{-2mm}
\vskip -5mm
\section{Future Work and Conclusions}
\label{sec:conclusion}
%% We have  defined a notion of  liveness on structured
%% data; this generalizes classical liveness and strong
%% liveness.    We  started   with   a  general   fully
%% context-sensitive analysis  which we  proved correct
%% with respect  to a minefield semantics  (this models
%% the  effect  of  garbage  collection  between  every
%% evaluation step).

@@

\appendix
\section{Soundness Result for Liveness Analysis under Minefield Semantics}
\label{sec:sound}
We first need an auxiliary  result about minefield semantics. Consider
a trace of  a minefield execution.  For every  minefield state $(\rho,
S,   \heap,  e(s),   \sigma)$   that   appears  on   the   LHS  of   a
$\rightsquigarrow$ step, the demand $\sigma$ on the expression $e$ (or
application $s$) is  non-null.  This can be proved by  an induction on
the number  of steps leading  to the  minefield state.  The  base step
holds   because  the   demand   $\sigma_{\mathit{all}}$   on  $(\mainpgm)$   is
non-null. For the inductive step we  observe that for each step of the
minefield semantics, if the demand $\sigma$  on the LHS of a minefield
step  is non-null,  the  demand  on the  RHS  is  a transformation  of
$\sigma$ (for example $(\clazy\cup\acar)\sigma$) which is also non-null.
